\documentclass[a5paper,11pt,titlepage]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{txfonts}
\usepackage[margin=0.75in]{geometry}
\usepackage{tcolorbox}

\title{Zen and the Art of Airbus Failure Handling}
\author{Jon Hurst}

\begin{document}
\maketitle

\begin{tcolorbox}[colframe=red]
\section*{Disclaimer}

Please clearly understand that there is \emph{nothing whatsoever}
official about this essay. I describe herein a mental model of failure
handling that I find useful. I publish it in the hope that, in part or
in whole, it may be of some use to others. Your mileage may assuredly
vary.

\end{tcolorbox}

\section{Model Description}

Picture, if you will, a mountain spring filling a small stream. This
spring represents your cognitive capacity.

A bamboo channel leads off the stream, with a sluice gate to control the
amount of water it drains. This channel represents the amount of
capacity that is being used for the primary tasks of flying of the
aircraft for PF and the monitoring of same by PM. It should \emph{never}
run dry!

Five large bowls sit by the stream. Mine are made of white marble, but
feel free to make yours out of whatever you wish. They are labeled:

\begin{enumerate}
\item Diagnosis
\item Containment
\item Strategy
\item Tactics
\item Review
\end{enumerate}

Finally, you have a ladle, which allows you to transfer water from the
stream to the bowls. Using this ladle represents the triage process.

When faced with a failure, your aim is to finish up with all the bowls
sufficiently full for a successful outcome. During the process of
filling the bowls the bamboo channel must always have sufficient
flow and the bowls must be, moment to moment, in a state such that the
immediate challenges are met.

\section{An example}
\label{sec:example}
So you are charging down the runway, you pass $V_1$, and you suddenly
find yourself needing a lot of rudder to keep straight. PM remarks
``engine one fail''. The sluice gate is wide open, as you are suddenly
dealing with non-standard handling occuring in a high workload flight
phase. There is just enough in the stream for a splash into the
Diagnosis bowl. This is just a recognition of symptoms. You diagnose
asymmetry and probable loss of thrust.

Triage says the next splash goes in the Tactics bowl. You deploy your
``asymmetric takeoff'' tactical unit. You mentally rehearsed this unit
just before applying thrust for takeoff, so the splash you can afford is
sufficient. You get safely airborne.

The Strategy bowl mustn't remain dry for long: your requirement is
``safe flight path'', your preference to minimise workload. Your options
are stay on the SID, fly the EOSID or fly an immediate VMC recovery. You
note that you are about to go IMC, so the EOSID it is. Luckily, this is
a STD EOSID, so you have a bit of breathing space; you just need to
climb straight ahead until you have accelerated to green dot. You tell
PM that this is the plan, and ask him to ``Pull Heading''. PM does so,
puts out a quick Mayday to tell ATC about the track deviation, asks them
to standby then makes an ``Attention Crew at Stations'' PA.

Next, another splash in the Tactics bowl to do something about the
amount of cognitive flow being diverted down the flying and monitoring
channels. You trim out the rudder and engage an autopilot. Suddenly
there is plentiful water in the cognitive stream.

So far, you've just been treating the worst symptoms; now that more
capacity is available, you add a good ladleful to the Diagnosis bowl.
You ask the PM to ``Confirm the failure''. PM runs through the
indications from the system pages and panels and offers up a diagnosis of
``engine 1 failure with damage''; you concur and tell them so.

Triage now says that getting the engine secured is the highest priority
item. You add a good ladleful to the Containment bowl. There are no OEBs
that affect, and the first ECAM procedure will secure the engine, so you
take the radios and ask for ``ECAM actions''. You successfully secure
the engine and ``Clear Eng''.

What now? The highest priority item is now the 10 minute TOGA limit. You
need to be clean before you can reduce the thrust, so you add another
ladleful to the Tactical bowl and apply the ``level acceleration and clean
up'' unit.

The strategic plan in force calls for a turn towards the EOSID holding
point at green dot at this point. However, you are now above MSA and the
holding point is 25 miles out to sea. You bring the Strategy bowl up to
the next level. Your apply ``head towards somewhere sensible'', options
are departure aerodrome or takeoff alternate. The weather was OK when
you left, so you choose to head to a late downwind position at the
departure aerodrome. You request an initial vector from ATC.

All immediate challenges are met, so back to the Containment. You
request ``Continue ECAM'' and continue with standard ECAM procedures
until ``ECAM actions complete''.

Time to top off the Strategy bowl. Requirements are a ``suitable
destination and a fallback option''. Preferences are for a proximate,
familiar airport with simple procedures, conservatively long runways and
reasonable weather without icing conditions. With PM you gather data,
carry out required calculations and generate options. From these options
you designate the departure airport as your primary and the takeoff
alternate as your fallback. You use the NITS format to inform ATC of
your plan, then request the ``SCCM to the flight deck''. You add a final
splash to the Diagnosis bowl by getting a report of symptoms detected
from the cabin, then inform the SCCM of the plan, again in the form of a
NITS brief. Finally, you make a quick, reassuring PA to the passengers.

Now it is time for the Review bowl. You critically visit each of the
other bowls in turn, methodically checking for correctness of actions
and absence of omissions.

That done, you top off the Tactics bowl. You plan modified anti-ice
procedures, fuel balancing procedures, modifications to windshear/EGPWS
procedures, asymmetric approach and landing and asymmetric
go-around. You set up appropriately, brief and fly a text-book
single-engine approach and landing.

You head to the bar.

\section{Discussion of model components}

\subsection{Flying and monitoring}

History has shown that it is perfectly possible to get so involved in
the handling of a failure that the flight path is fatally
disregarded. Sufficient cognitive capacity must always be devoted to
the flying task by PF and to the monitoring task by PM. Active awareness
of how difficult it is to do this in the face of a complex failure is
the only real defence.

Since the flying/monitoring task is taking resource from a finite pool,
any steps that reduce the amount of capacity required for sufficiency
are helpful.

The main step that can be taken is to grab any safe opportunity to train
manual handling skills. It is an unfortunate fact that the more severe a
failure is, the more chance there is that you will be flying the
aircraft manually, possibly with degraded handling characteristics. If
you have been regularly honing your skills on fine sunny days, the gap
that must be filled when those skills are required on a dark stormy
night will be much smaller, and the amount of spare capacity available
for handling the underlying failure will be much greater. This applies
to monitoring as much as to handling. The automatics are generally so
reliable that atrophied monitoring skills can go unpunished for a very
long time; regularly acting as PM when PF is handling manually provides
a good antidote.

If it is available, use of the autopilot makes an enormous
difference. In particular, monitoring the autopilot flying a hold in VMC
requires a mere trickle of cognitive capacity; IMC when above MSA
requires a little more, but is still a good place to be. If this hold
happens to be at the end of the downwind leg to a suitable runway, then
a lot of your problems are already solved.

If the autopilot is not available, there is often a significant
asymmetry between the capacity drain for handling and the capacity drain
for monitoring, particularly if the handling is being done well. This
can easily lead to PF not having the capacity to properly monitor PM's
actions or properly absorb the intricacies of the failure. Temporarily
swapping control is useful in these circumstances, giving PF the
capacity to review what has been done and expand their situational
awareness bubble.

\subsection{Triage}
The triage process is visualised as a ladle. This is related to the
concept of chunking of cognitive capacity in the face of a requirement
to simultaneously carry out multiple novel tasks.

Human beings are not particularly good at real multi-tasking; we
actually rely on motor programs and rapid task switching. If a task is
novel, we will not have a suitable motor program unless we can re-task
something that is not novel.\footnote{Re-tasking motor programs is often
  a good technique. For example, for a go-around everything after
  reading the FMAs is the same as a flap 3 takeoff; it is worth
  practising a flap 3 takeoff from time to time so you can have the
  motor program available for the go-around.} Interrupting a task before
it is complete often leads to errors, so switching tasks too rapidly
should be avoided. Equally, not switching tasks rapidly enough leads to
``no-one flying the plane'' scenarios. The trick, then, is to
methodically break workflow down into short, focused, completable
chunks, each to be applied at the most opportune moment. The moment to
moment selection of the optimum chunk to apply is the process of triage.

The efficient application of triage relies heavily on having a suitably
stocked ``bag of tricks''. There is, unfortunately, no shortcut to
developing this. It comes from systematic study of the abnormal
procedures and techniques set down by Airbus in the QRH, FCTM and FCOM,
carried out with an eye for commonalities and adaptions of normal
handling techniques. It is not necessary (and likely not possible) to
memorise every checklist and procedure, but it should be possible to
give a quick summary of what you would likely do for any given symptom
and use the checklists and ECAM to fill in the gaps.

\subsection{Diagnosis}

It is vital to remember that, at best, the ECAM provides a data point
for a diagnosis, not the diagnosis itself. There are scenarios such as
low engine oil pressure where different sensors provide data to the ECAM
and system pages, and a faulty sensor should be assumed in the event of
disagreement. There are scenarios such as triple ADR failure where the
ECAM will simply refuse to believe that such a thing could happen and
will send you round in circles. There are scenarios such as unreliable
airspeed that the ECAM will simply lie to you (noisily) about what is
happening.

Diagnosis will generally be easier in the simulator than in real
life. In the simulator a programmer has carefully arranged for all the
symptoms to appear as they should. Real life is a lot messier. I have
had four unconnected systems fail simultaneously, and it took twelve
days in the hanger to find the offending burnt through wiring loom. It
will not always be possible to settle on a root cause, in which case you
will have to work with an incomplete diagnosis and settle for treating
the symptoms; this is an extremely uncomfortable position to find
yourself in.

Experience and good technical knowledge are the two pillars of accurate
diagnosis. It is always preferable to gain the benefit of someone else's
bad day out rather than your own, so reading incident reports is
definitely a good idea. Beyond that there is unfortunately no real
substitute for time spent reading the FCOM and FCTM.

\subsection{Containment}

The best containment outcome is to recover the failed system. Many
Airbus containment procedures thus amount to ``Turn it off, turn it back
on, if it still doesn't work, turn it off again.'' Sometimes it is a
case of ``wait until it has cooled down'' or ``wait until at a lower
altitude'', but the general principal holds.

QRH and ECAM procedures often mix in strategic and tactical
recommendations and it is easy to get drawn into detailed planning at
inopportune times as a result. A good example is Flaps/Slats faults. The
basic containment provided by ECAM is utterly trivial: work out the
limiting speed for the lift devices in their current position and fly
about 10 knots slower than that, then attempt to recycle the devices if
they are not locked. As soon as you call for the QRH procedure, however,
you will be drawn into a whole bunch of techniques, notes, tables and
conditional subsections that are not going to improve your immediate
situation. It is often useful to simply acknowledge the existence of
such recommendations and agree to defer detailed examination of them for
a later planning phase when more capacity is available.

In general, it is extremely useful to have studied all the published
abnormal procedures in the FCOM and QRH so that you have a pretty good
idea about how you will go about containing any given failure. The
ECAM/QRH checklists then become a tool for polishing your response
rather than a primary information source. Airbus abnormal procedures are
sometimes not a model of clarity, and encountering a checklist for the
very first time during a failure scenario may well lead to confusion and
errors.

\subsection{Strategy}

Strategy is the big picture stuff. It is strongly informed by the
diagnosis and effectiveness of containment.

Strategic planning generally rises through three levels, with increasing
amount of cognitive capacity required for each:

\begin{enumerate}
\item Establish a safe flight path
\item Head in a sensible direction
\item Specify destination and fallback position
\end{enumerate}

Each stage of strategic planning can be broken down into four
necessarily sequential steps:

\begin{enumerate}
\item Declare requirements and preferences
\item Generate options
\item Choose a primary and fallback option
\item Communicate the plan
\end{enumerate}

A discussion and clear declaration of requirements and preferences is a
great barrier to rushed and faulty strategic decision making.

The usefulness of the MEL in requirement generation should not be
underestimated. If Airbus allow you to dispatch with something, then,
provided you have the fuel and any MEL conditions for that dispatch are
met, continuing to scheduled destination is probably the right move. The
MEL can also generate ideas for additional containment and tactical
procedures. A classic example of this is the N/W STRG FAULT, which
becomes a lot less frightening once you realise that you can (just)
dispatch with it, and that there are a heap of good suggestions in
the associated MEL OPS procedure.

On the flip side, if the MEL forbids dispatch with a failure, a decision
to divert should be strongly considered. An example of this is the
relatively common Blue System Electric Pump failure which feels fairly
minor (mainly just loss of a spoiler per side and slow slats, with the
RAT available if another hydraulic system is lost), yet Airbus won't let
you dispatch with it. This is an indication that the redundancy provided
by that pump is designated critical by Airbus and thus pushing on to
a distant destination might be considered overly brave.

There are usually three sets of people outside the cockpit that need to
know your plans: ATC, the cabin crew and the passengers. Under stress,
non-standard communication can easily become muddled. A good tool to
ensure structure is the NITS brief, which can be used to clearly
disseminate the plan to all three groups:

\begin{itemize}
\item \textbf{N}ature of failure
\item \textbf{I}ntentions
\item \textbf{T}ime
\item \textbf{S}pecial requirements
\end{itemize}

\subsection{Tactics}

Tactics are the detailed implementation required to achieve your
strategic goals. They are a ``bag of tricks'' containing supplementary
flying, automation and systems handling techniques.

A tactic is most effective when it is constructed as a rapidly
completable block of simple, logically connected actions that can be
considered as a single unit. For example, if you have a well rehearsed
``asymmetric take-off'' tactic\footnote{Mine is slower rotation,
  $12\frac{1}{2}^\circ$ pitch target becoming a $V_2$--$V_2+15$ airspeed
  target, max bank angle $15^\circ$, TOGA when flight mode blended} that
you can reliably execute as a unit, the initial part of an EFATO becomes
trivial. Similarly you can break down the RTO into an ``RTO initiation''
tactic and an ``RTO cleanup'' tactic, with a consideration of fire
versus wind between them. When you start breaking things down in this
way, even complex failures such as dual hydraulic failures end up
boiling down to a short unthreatening list of common tactical
units. Furthermore, you end up with a flexible toolkit for dealing with
failures and combinations of failures that Airbus has not planned
for. The difficult part, of course, is stocking your ``bag of tricks''
in the first place; this can only be achieved by studying the manuals
with an eye for common tactical units.

From a tactical point of view, failure scenarios tend to be made up of
two phases, starting with a ``catch up'' phase where tactics are
inserted individually on an ad hoc basis to meet immediate challenges
and evolving into a ``tactical stack'' phase where a pre-planned series
of tactics are interleaved with the standard operation. For the latter
it can be very helpful to divide the stack into six flight phases:
approach, landing, stopping, parking, go-around and diversion. Write out
a bulletted list of the tactics you intend to deploy so that you can
easily check they have been actioned once the workload starts to build.

\subsection{Review}

It was trying to find a solution to the problem of ineffective review
that was the seed for this mental model. It is easy to think that
turning to the other pilot and saying ``have we forgotten anything''
achieves something useful. On every occasion that I did this, we failed
to come up with anything, even on those occasions we had indeed
forgotten something.

Breaking the process into Diagnosis, Containment, Strategy and Tactics
allows a much more methodical approach, with each area examined in turn.
To review Diagnosis, you review the systems pages and panels looking for
contradictory evidence. To review Containment you use the RCL button to
check through the cleared ECAM procedures, check through QRH procedures
for errors and omissions (particularly if you were not the one doing
them), consider any available resets that may not have been tried and
refer to extended checklists and the MEL as appropriate. To review
Strategy consider whether requirements and preferences remain the same
and whether the chosen option and fallback position remain optimum. To
review Tactics you check through the tactical stack with an eye for
omissions.

The optimum quantity and timing of Reviews is a non-trivial question
that needs to be adapted to the situation. It is easy to end up burning
time and capacity pointlessly re-reviewing, so my default is to carry
out a full review of Diagnosis, Containment and Strategy once the final
Strategic planning is in place, then to assume that unless something
drastic happens, those areas are solid. Then, at a suitably low
work-load time on the approach I will use the bulletted list of tactical
units I created to ensure that everything that was planned to have been
done up to that point has been completed and any remaining items are at
the front of my mind.

\section{Compatibility}
\subsection{DODAR}

DODAR is easyJet's recommended failure management process. It is an
acronym standing for \textbf{D}iagnosis; \textbf{O}ptions;
\textbf{D}ecide; \textbf{A}ssign tasks and Action; \textbf{R}eview
Decision.

Each of the individual items of DODAR is useful and appropriate, but I
find it inflexible and incomplete when applied as a ``process''. All the
elements of DODAR are thus included in the model, but each element is
evolved through multiple levels with the timing driven by triage. This
is arguably just a case of being more explicit about how DODAR can be
applied in practice; if you accept this then the model can be considered
fully compatible with DODAR.

\subsection{Simulator scripts}

There are a few sections of recurrent simulator tests, such as the
EFATO, that are often examined in the form of scripts such that details
down to the level of the order that buttons are pressed become
significant.

It may be noted that the EFATO example in section~\ref{sec:example} is
indistinguishable from the EFATO simulator script provided that the
relevant triage decisions are pre-determined and that the procedures you
pull out of your ``bag of tricks'' are the same at button order
level. The same is true of all of the scripted sections.

Thus, this model can easily be made compatible with the scripted
sections of simulator testing by making sure your procedure chunks are
compatible and learning the scripts as a short series of triage
decisions. The procedure chunks will then be available for re-use and
adaption in less well defined events, even if it won't do much for
improving your triage process itself.

\end{document}
