\documentclass[a5paper,11pt,twoside]{book}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\usepackage[left=0.75in, bottom=1in]{geometry}
\usepackage{microtype}
\usepackage{color}
\usepackage[color]{changebar}
\cbcolor{blue}
\usepackage{array}

\title{A320 Family Non-Normal-Notes\\~\\\large{Version 2.6}}
\author{Jon Hurst}
\date{}
\newcommand{\multicite}[1]{
  \nopagebreak
  \noindent{\footnotesize\color{blue}{[ #1 ]}}
}

\newcommand{\V}[1]{V\textsubscript{#1}}

\begin{document}
\maketitle
\tableofcontents

\chapter{Operating techniques}
\section{Rejected Takeoff}

The decision to reject rests solely with CM1. This decision is communicated with
the words “Stop” or “Go”. “Stop” implies that CM1 is taking control of the
aircraft. Below 100kt the RTO is relatively risk free and a decision to stop
should be made for any ECAM and most other problems. Above 100kt the RTO may be
hazardous and stopping should only be considered for loss of engine thrust, any
fire warning, any uninhibited ECAM\footnote{A change was introduced in Feb 2018
  whereby five amber ECAM cautions requiring a high speed stop are specifically
  listed. An exhaustive search of amber ECAM cautions that are not inhibited
  between 80kt and lift off revealed only two cautions in addition to this list:
  ENG 1(2) THR LEVER DISAGREE if the FADEC automatically selects idle thrust and
  FWS FWC 1+2 FAULT. The first of these should theoretically never happen due to
  FADEC logic. The second generates a message on the EWD but no master caution,
  since it is the computers that generate master cautions that have failed. You
  could therefore modify this rule to: stop for any ECAM warning or caution
  except the caution-like FWS FWC 1+2 FAULT.} or anything which indicates the
aircraft will be unsafe or unable to fly.

If a stop is required, CM1 calls “Stop” while simultaneously bringing the thrust
levers to idle, then to max reverse. If the stop was commenced below 72kt the
ground spoilers will not automatically deploy and the autobrake will therefore
not engage. Monitor automatic braking, and if there is any doubt, apply manual
braking as required. If normal braking fails, announce “Loss of braking” and
proceed with the loss of braking memory items (see <xref
linkend="loss-of-braking"/>). If the reason for the stop was an engine fire on
the upwind side, consider turning the aircraft to keep the fire away from the
fuselage. If there is any chance of requiring evacuation, bring the aircraft to
a complete halt, stow the reversers, apply the parking brake, and order
“Attention, crew at stations” on the PA. If evacuation will definitely not be
required, once the aircraft’s safety is assured the RTO can be discontinued and
the runway cleared. In this case make a PA of “Cabin crew, normal
operations”.

During this initial phase, CM2 confirms reverse (“Reverse green”), confirms
deceleration (“Decel”), cancels any audio warnings, informs ATC and announces
“70 knots” when appropriate. CM2 then locates the emergency evacuation
checklist.

Once the aircraft has stopped, CM1 takes the radios and asks CM2 to carry out
any required ECAM actions. Whilst the ECAM actions are being completed, CM1 will
build up a decision as to whether to evacuate. If an evacuation is required see
<xref linkend="evacuation"/>. Otherwise order “Cabin crew, normal operations”.

If the aircraft has come to a complete halt using autobrake MAX, the brakes can
be released by disarming the spoilers.

If, following an RTO, a new takeoff is to be attempted, reset both FDs, set the
FCU, then restart SOPs from the After Start checklist. Carefully consider brake
temperatures; temperature indications continue to climb for some time after a
significant braking event.

\multicite{EOMB 3.10, FCTM PRO.AEP.MISC}


\section{Failures during takeoff when above \V{1}}

If an engine has lost thrust, apply rudder conventionally. At V\textsubscript{r}
rotate towards an initial pitch target of 12½° at a slightly reduced rate, then
target speed \V{2} to \V{2}+15kt. Bank angle should be limited to 15° when more
than 3kt below manoeuvring speed for the current configuration\footnote{This is
a conservative rule of thumb. If the FMGC has correctly identified an engine out
condition, the FD/AP will automatically limit bank angle according to a less
conservative algorithm. [FCOM SYS.22.20.60.40]}.

When the ground to flight mode transition is complete\footnote{Introducing TOGA
during the ground to flight mode transition (commences as the pitch increases
through 8°, complete after 5 seconds) results in a pitch up moment at a time
where the effect of stick pitch control is not wholly predictable: the stick
will need to be moved forward of neutral to counteract the introduced pitch
moment and then returned to neutral as flight mode blends in. A slight pause
before selecting TOGA results in much more normal and predictable handling.},
select TOGA (FLX may be used but this tends to allow speed to decay unless pitch
is reduced), adjust and trim rudder to maintain β target and request “pull
heading”. If the EOSID follows the track of the cleared SID, NAV may be used,
but this is very rare with easyJet EOSIDs.  Engage the autopilot once gear is up
and rudder is trimmed.

Whilst below 400ft, the only failure related actions should be:
\begin{itemize}
\item If applicable, PM should state the title of the first displayed ECAM
  procedure, but should delay confirmation.\footnote{ The Feb 2018 update to
  EOM-B removed the “engine failure” and “engine fire” calls, but leaves the
  timing of the reading of the ECAM title ambiguous. The advice given here is
  one interpretation. Since these calls were good for situational awareness,
  replacing them with a bald statement of the first ECAM procedure title seems
  sensible.}

\item Cancellation of master warning or master caution when both pilots confirm
  they are aware of it.

\item Heightened awareness of the possibility of missing essential normal
  actions, such as calling rotate or raising the gear due to the distraction of
  the failure.
\end{itemize}


Once above 400ft with safe flight path assured, decide on an initial
strategy. In general, where a loss of thrust has occurred or is anticipated, the
strategy will be to fly the EOSID with a level acceleration segment (see <xref
linkend="eosid"/>). Otherwise, it will be to remain on the normal SID and fly a
normal climb profile. Any deviation from the cleared SID will require ATC to be
informed as a priority, usually as part of a PAN or MAYDAY message. In rare
cases where the cleared SID requires a very early turn it may be necessary to
determine and action a strategy when below 400ft. If this is the case, it must
be thoroughly briefed.

Once the flight path strategy has been agreed and actioned, the failure can be
triaged, diagnosed and contained. Once the diagnosis is agreed, PF will take the
radios and request PNF to carry out the containment actions. The standard
phraseology is “My Radios, ECAM actions”. PF taking the radios is also a good
trigger to consider a Mayday and an “Attention crew at stations”.

When applying ECAM procedures, PF is responsible for moving the thrust levers
once confirmed by PNF. PNF is responsible for everything else, but movement of
engine master switches, IR selectors and any guarded switch must be confirmed
with PF.

PNF indicates high priority tasks are completed with the phrase “Engine is
secure.”  This call is not official SOP (i.e. it is not mentioned in any
manual), but has evolved to be standard in the sim since it is PF’s trigger to
interrupt ECAM and accelerate.  High priority tasks are defined as:

\begin{itemize}
\item For engine failure, the master switch of the affected engine has been
  turned off.

\item For engine fire, one squib has been fired and the fire warning has
  extinguished or both squibs have been fired.
\end{itemize}

\multicite{FCTM PRO.AEP.ENG}




\section{EOSID}

Before the divergence point (the last common point between the SID and the
EOSID), if the aircraft detects a loss of thrust the EOSID will be displayed as
a temporary flight plan. In this case the temporary flight plan can be inserted
and NAV mode used. Otherwise it will be necessary to pull heading and manually
follow either the yellow line or bring up a pre-prepared secondary flight plan
and follow the white line.

If beyond the divergence point, pull heading and make an \emph{immediate} turn
the shortest way onto the EOSID. Airbus specifically recommends against this in
FCOM.AS.22\_20.60.40, but easyJet states it as policy in EOMB 4.4.4.

Electing to fly the EOSID implies a level acceleration segment:

\begin{itemize}
\item Initially fly a TOGA climb at the higher of \V{2} or current speed, up to
  a limit of \V{2}+15kt. If a FLEX takeoff was carried out, a FLEX climb is
  permissible. This climb is continued until all high priority tasks are
  complete (see <xref linkend="efato"/>) and the aircraft is above single engine
  acceleration altitude (usually 1000ft QFE, but may be higher if so specified
  by the take-off performance calculation). If the FMGS has detected the engine
  out condition, the automatic mode change from SRS to CLB will be inhibited; if
  not, intervention with selected modes will be required to prevent untimely
  acceleration.

\item The next segment is a TOGA level acceleration and clean up, either to Conf
  1 and S speed if an immediate VMC return is desired or to Conf 0 and green
  dot. Again FLEX may be used if a FLEX takeoff was carried out. Level
  acceleration is usually achieved by pushing V/S; if the FMGS has detected the
  engine out condition, all preselected speeds entered in the MCDU will have
  been deleted, so the managed target speed should automatically move to
  250kt. The phrases “Stop ECAM” and “Continue ECAM” can be used to interrupt
  ECAM procedures in order to initiate this segment.

\item The final segment is a MCT climb segment to MSA, either at S speed if in
  Conf 1 or at green dot speed if in Conf 0. This is usually achieved in open
  climb; if the FMGS has detected the engine out condition, the managed target
  speed becomes dependent on flight phase, and in this case should automatically
  select green dot.
\end{itemize}

TOGA may be used for a maximum of 10 minutes.

If an EOSID is annotated as “STD”, then acceleration to green dot should be
completed prior to commencing the first turn. If “NON-STD”, the turn takes
priority.

\multicite{EOMB 4.4.4, FCOM DSC.22\_20.60.40}

\chapter{Miscellaneous}

\section{Emergency descent \emph{(memory item)}}

If an emergency descent is required, the Captain should consider taking control
if not already PF. PF initiates the memory items by announcing “Emergency
Descent.”

Don oxygen masks and establish communication.

PF then flies the emergency descent.  Descent with autopilot and autothrust
engaged is preferred. The configuration is thrust idle, full speed brake and
maximum appropriate speed, taking into account possible structural
damage.\footnote{According to Airbus, structural damage may be suspected if
there has been a “loud bang” or there is a high cabin vertical speed. When
limiting descent speed due to suspected structural damage, it is IAS rather than
Mach that is relevant.} Target altitude is FL100 or MORA if this is higher. If
speed is low, allow speed to increase before deploying full speedbrake to
prevent activation of the angle of attack protection. Landing gear may be used,
but speed must be below \V{LO} when it is extended and remain below \V{LE}. If
on an airway, consider turning 90° to the left.

PM’s only memory action is to turn the seatbelt signs on.\footnote{Prior to a
2017 update from Airbus, PM would complete a fairly lengthy list of memory items
at the start of the Emergency Descent procedure. It was found that PM was more
usefully employed monitoring PF’s actions, and hence most of these memory items
were removed to a read and do checklist to be completed once descent had been
initiated.} Initially, PM’s main task is to ensure that PF has promptly and
correctly initiated the descent.

Once the memory actions are complete and the aircraft is descending, PF should
call for the Emergency Descent Checklist (“My radios, Emergency Descent
Checklist”). This will lead PF to finesse the speed and altitude targets and
inform ATC of the descent; PM to set continuous ignition on the engines and set
7700 on the transponder. Both pilots then set their oxygen flows to the N
position\footnote{There may be insufficient oxygen to cover the entire emergency
descent profile if the oxygen masks are left set to 100\%.} and, if cabin
altitude will exceed 14,000ft, PM deploys the cabin oxygen masks. On easyJet
aircraft, the CIDS/PRAM will automatically play a suitable PA, so it is not
necessary for the flight crew to carry out the EMER DESCENT (PA) action.

Once level, restore the aircraft to a normal configuration. When safe to do so,
advise cabin crew and passengers that it is safe to remove their masks. To
deactivate the mask microphone and switch off the oxygen flow, close the oxygen
mask stowage compartment and press the “PRESS TO RESET” oxygen control slide.

\multicite{EOMB 3.80.2, QRH AEP.MISC, FCOM AEP.MISC, FCTM AEP.MISC}


\section{Windshear \emph{(memory item)}}

\subsection{Reactive}

The windshear detection system is a function of the Flight Augmentation Computer
(FAC). It only operates during the takeoff and landing phases with at least CONF
1 selected. In the takeoff phase, warnings are provided from 3 seconds after
lift off until 1300ft RA is achieved. In the landing phase warnings are provided
between 1300ft RA and 50ft RA. A warning is indicated by a red “WINDSHEAR” flag
on the PFD and a “WINDSHEAR, WINDSHEAR, WINDSHEAR” aural warning.

 When on the ground, windshear is \emph{only} indicated by significant airspeed
 variations. It is possible that these fluctuations may cause \V{1} to occur
 significantly later in the takeoff run then it should. It therefore falls to
 the Captain to make an assessment of whether sufficient runway remains to
 reject the takeoff, or whether getting airborne below \V{r} would be the better
 option. If the takeoff is to be continued in windshear conditions, call
 “Windshear, TOGA” and apply TOGA power.  Rotate at \V{r} or with sufficient
 runway remaining and follow SRS orders.\footnote{“Sufficient runway remaining”
 is actually Boeing advice – Airbus offers no guidance for the case where there
 is insufficient runway available to stop nor to rotate at normal speeds.} SRS
 will maintain a minimum rate of climb, even if airspeed must be sacrificed.

If a warning occurs when airborne, call “Windshear, TOGA”, apply TOGA power and
maintain current configuration. The autopilot can fly the escape manoeuvre as
long as $\mathrm{\alpha_{req}<\alpha_{prot}}$. If the
autopilot is not engaged, follow the SRS orders on the FD. If the FD is not
available, initially pitch up to 17.5°, then increase as required.

Do not change configuration until out of the windshear.

In severe windshear, it is possible that ALPHA FLOOR protection will
activate. As TOGA will already be selected, this will have no immediate
effect. Once clear of the windshear, however, TOGA LK will be active, requiring
the autothrust to be disconnected to avoid an overspeed. This is most naturally
achieved by pressing the “instinctive disconnect pb” on the thrust levers then
flying manual thrust until the situation has sufficiently stabilised to
re-engage the autothrust.

\multicite{FCOM PRO.AEP.SURV}

\subsection{Predictive}

When below 2300ft AGL, the weather radar scans a 5nm radius 60° arc ahead of the
aircraft for returns indicating potential windshear.

Alerts are categorised as advisory, caution or warning, in increasing order of
severity. Severity is determined by range, position and phase of flight. Alerts
are only provided when between 50ft and 1500ft, or on the ground when below
100kt.

All types of alert produce an indication of windshear position on the ND,
providing the ND range is set to 10nm. A message on the ND instructs the crew to
change range to 10nm if not already set. Cautions also give an amber “W/S AHEAD”
message on both PFDs and an aural “MONITOR RADAR DISPLAY” warning. Warnings give
a red “W/S AHEAD” message on the PFDs and either a “WINDSHEAR AHEAD, WINDSHEAR
AHEAD” or “GO AROUND, WINDSHEAR AHEAD” aural message.

If a warning alert occurs during the takeoff roll, reject the takeoff. If it
occurs during initial climb, call “Windshear, TOGA”, apply TOGA thrust and
follow SRS orders. Configuration may be changed as long as the windshear is not
entered.

If a warning alert occurs during approach, fly a normal go-around. If positive
verification is made that no hazard exists, providing that the reactive
windshear is serviceable the crew may downgrade the warning to a caution. If a
caution alert occurs during approach, consider use of CONF 3 and increasing
\V{APP} to a maximum of \V{LS}+15.

\multicite{FCOM PRO.AES.SURV}

\section{Unreliable airspeed \emph{(memory item)}}

Unreliable airspeed indications may result from radome damage and/or
unserviceable probes or ports. Altitude indications may also be erroneous if
static probes are affected.

The FMGCs normally reject erroneous ADR data by isolating a single source that
has significant differences to the other two sources. It is possible that a
single remaining good source may be rejected if the other two sources are
erroneous in a sufficiently similar way. In this case, it falls to the pilots to
identify and turn off the erroneous sources to recover good data.

The first problem is recognition of a failure, since the aircraft systems may be
unable to warn of a problem. The primary method of doing this is correlation of
aircraft attitude and thrust to displayed performance. Correlation of radio
altimeter and GPIRS derived data (available on GPS MONITOR page) may also aid
identification. The stall warning (available in alternate or direct law) is
based on alpha probes, so will likely be valid. Other clues may include
fluctuations in readings, abnormal behaviour of the automatics, high speed
buffet or low aerodynamic noise.

If the aircraft flight path is in doubt, disconnect the automatics and fly the
following short term attitude and thrust settings to initiate a climb:

\medskip
\begin{tabular}{|l|c|c|}
  \hline
  \textbf{Condition} & \textbf{Thrust} & \textbf{Pitch}\\\hline
  Below Thrust Reduction Altitude & TOGA & 15°\\\hline
  Below FL100 & CLB & 10°\\\hline
  Above FL100 & CLB & 5°\\\hline
\end{tabular}
\medskip

If configured CONF Full, select CONF 3, otherwise flap/slat configuration should
be maintained. The gear and speedbrake should be retracted. If there is any
doubt over the validity of altitude information, the FPV must be disregarded. If
altitude information is definitely good, the FPV may be used.

It is important to understand that at this stage, while the pilot has identified
that airspeed is unreliable, the aircraft systems have not. Thus flight envelope
protections based on airspeed data from unreliable ADRs may activate. This may
lead to pitch inputs from the flight computers that cannot be overridden with
the sidesticks. In this case, immediately switch off any two ADRs; this causes
the flight computers to revert to Alternate Law with no protections, and thus
allows control of the aircraft to be regained.

Once the flight path is under control and a safe altitude is attained, the
aircraft should be transitioned into level flight. Refer to the QRH to
extract a ballpark thrust setting, a reference attitude and a reference speed
for the current configuration, bearing in mind that an auto-retraction of the
flap may have occurred. Set the ballpark thrust setting and adjust pitch
attitude to fly level; if barometric altitude data is considered accurate use
the VSI, otherwise fly a constant GPS altitude. The thrust should then be
adjusted until level flight is achieved with the reference attitude. Note that
in the radome damage case, the required N1 may be as much as 5\% greater than
the ballpark figure. Once stable, the speed will be equal to the reference
speed.

If there is insufficient data available to fly level (e.g. GPS data unavailable
and barometric data unreliable), fly the reference attitude with the ballpark
thrust setting. This will give approximately level flight at approximately
reference speed.

With the speed now known, the ADRs can be checked to see if any are giving
accurate data. If at least one ADR is reliable, turn off the faulty ADRs. GPS
and IRS ground speeds may also be used for an approximate cross check.

If all ADRs are considered unreliable, turn off any two of them; one is kept on
to provide stall warning from the alpha probes. More recent aircraft have backup
speed/altitude scales based on AOA probes and GPS altitudes which are activated
when below FL250 by turning off the third ADR. The ALL ADR OFF procedure in
QRH AEP.NAV describes the use of these scales, but it boils down to fly the
green on the speed scale and anticipate slightly reduced accuracy from the
altitude scale.  For aircraft without this functionality, tables are provided in
section AEP.NAV.USI of the QRH to enable all phases of flight to be flown using
just pitch and thrust settings. Acceleration and clean up are carried out in
level flight. Flap 1 can be selected as soon as climb thrust is selected, flap 0
once the appropriate S speed pitch attitude from the table on the first page of
the QRH AEP.NAV.USI procedure is reached. Configuration for approach is also
carried out in level flight, stabilising in each configuration using the
technique described above. The approach is flown in CONF 3 at an attitude that
should result in VLS+10 when flying a 3° glide. Landing distance will be
increased.

\multicite{QRH AEP.NAV, FCOM PRO.AEP.NAV, FCTM PRO.AEP.NAV}

\section{Incapacitation}

Take control, using the stick priority button if necessary. Contact cabin crew ASAP. They
should strap the incapacitated pilot to his seat, move the seat back, then recline it. If there
are two cabin crew available, the body can be moved. Medical help should be sought from
passengers, and the presence of any type rated company pilots on board ascertained.

\multicite{FCTM PRO.AEP.MISC}

\section{Forced Landing (inc. Ditching)}

There are two scenarios where off field forced landing or ditching would be
considered: either you have insufficient energy to reach a suitable airfield
(e.g actual or impending fuel exhaustion, catastrophic failure of both engines),
or you have insufficient time to do so (e.g. uncontained fire).

Ditching and off field forced landing without power are discussed in <xref
linkend="all-engine-failure"/>. Support for such landings is provided by the ALL
ENGINES FAILURE ECAM and the ALL ENG FAIL QRH checklist.

Support for ditching and forced landing with power is provided by the QRH
Ditching and QRH Forced Landing checklists respectively. Necessarily implicit in
these checklists is the assumption that the aircraft is fully servicable, which
is unlikely to be the case. There is also an assumption that plenty of time is
available for extensive preparation of cabin and cockpit. It is highly likely
that these checklists will need adapting to the situation.

The fundamentals are the same with or without power. Ditching will be gear up
with a target pitch attitude of 11° and minimal vertical speed, and should be
made parallel to the swell unless there are strong crosswinds, in which case an
into wind landing is preferred. Forced landing will be gear down with the
spoilers armed. The aircraft should be depressurised for the landing, with the
Ditching PB pushed for the ditching case; in the without power cases this is
achieved using RAM AIR, and this is the same for with power forced landing. The
“with power” ditching achieves depressurisation by turning off all bleeds, which
provides a slightly more watertight hull.

The main difference between the with and without power cases is that max
available slats and flaps are used in the former case, wheras CONF 2 is mandated
for the latter. Approach speeds must also be high enough to prevent RAT stall
(i.e. >140kt) if it is being relied upon. The combination of these factors means
that much lower landing speeds can be achieved if power is available.

\multicite{QRH AEP.MISC}

\section{Evacuation}

Evacuation should be carried out in accordance with the emergency evacuation
checklist. The easyJet procedure is for CM1 to call for the checklist and then
send a Mayday message to ATC before commencing the checklist.

The first two items confirm the RTO actions of stopping the aircraft, setting
the parking brake and alerting the cabin crew. The next item confirms ATC has
been alerted.

The next four items prepare the aircraft for evacuation. If manual cabin
pressure has been used, CM2 checks cabin diff is zero, and if necessary manually
opens the outflow valve. CM2 then shuts the engines down with their master
switches, and pushes all the fire buttons (including the APU). Confirmation is
not required before carrying out these actions.\footnote{The Airbus Training
Study Guide was recently (mid 2017) changed to align with the FCTM.}  In
response to the next checklist item, “Agents”, CM1 decides if any extinguishing
agents should be discharged and instructs CM2 to discharge them as
required. Engine agent 2 will not be available. Agents should only be discharged
if there are positive signs of fire.

Finally, order the evacuation. This is primarily done with the PA “Evacuate,
unfasten your seat belts and get out”, with the evacuation alarm being triggered
as a backup.

\multicite{EOMB 3.80.1, FCOM PRO.AER.MISC, FCTM PRO.AER.MISC}


\section{Overweight landing}

A landing can be made at any weight, providing sufficient landing distance is
available. In general, automatic landings are only certified up to MLW, but the
FCOM specifies that, for the A319 only, autoland is available up to 69000kg in
case of emergency. The preferred landing configuration is CONF FULL, but lower
settings may be used if required by QRH/ECAM procedures. QRH AER.MISC.OL also
specifies CONF 3 if the aircraft weight exceeds the CONF 3 go around limit; this
will only ever be a factor for airfields with elevations above 1000ft. Packs
should be turned off to provide additional go around thrust. If planned landing
configuration is less than FULL, use 1+F for go-around.

It is possible that S speed will be higher than \V{FE next} for CONF 2. In this
case, a speed below \V{FE next} should be selected until CONF 2 is achieved,
then managed speed can be re-engaged.

In the final stages of the approach, reduce speed to achieve \V{LS} at runway
threshold. Land as smoothly as possible, and apply max reverse as soon as the
main gear touches down. Maximum braking can be used after nosewheel
touchdown. After landing, switch on the brake fans and monitor brake
temperatures carefully. If temperatures exceed 800°C, tyre deflation may occur.

\multicite{QRH AER.MISC, FCOM PRO.AER.MISC, FCTM PRO.AER.MISC}

\section{Engine failure in cruise}

Engine out ceiling is highly dependent on weight; ISA deviation also has a
modest effect. It will generally lie between FL180 and FL350.

The first action will be to select both thrust levers to MCT so as to allow the
autothrust its full engine out range. If the N1 gauges indicate a thrust margin
exists, then the aircraft is below engine out ceiling; descent may be
appropriate to increase the available thrust margin, but there is no immediate
threat. If, however, the N1 gauges indicate that the autothrust is commanding
MCT, and the speed is still decaying, then the aircraft is above engine out
ceiling and prompt execution of a drift down procedure is required.

Drift down with autopilot engaged in OP DES is preferred. Engagement of this
vertical mode normally results in the autothrust commanding idle thrust, which
is not what is desired. Thus, having set the thrust lever to MCT, the autothrust
is disconnected. The PROG page provides a REC MAX EO flight level to use as an
altitude target. If the speed decay is modest, it may be possible to alert ATC
before initiating the descent, but in-service events have shown that speed decay
is often very rapid, requiring descent initiation to be prioritised.

Once drift down has been initiated, a decision needs to be made about speed. If
obstacles are a concern, the lowest drift down rate and highest ceiling are
achieved at green dot. Airbus refers to drifting down at green dot as “Obstacle
strategy”. Flying at green dot reduces the chance of the FADECs automatically
relighting the failed engine as the engine will be windmilling more
slowly. Therefore, if obstacles are not a concern, M.78/300kt is flown, a speed
that will always fall within the stabilized windmill engine relight envelope;
Airbus refers to this as “Standard Strategy”.

If obstacles remain a problem, MCT and green dot speed can be maintained to give
a shallow climbing profile. Once obstacles are no longer a problem, descend to
LRC ceiling (use V/S if $<$500 fpm descent rate), engage the autothrust and
continue at LRC speed.

\multicite{FCTM PRO.AEP.ENG.EFDC}
\section{Single engine circling}

It may not be possible to fly level in the standard circling configuration of
CONF 3 gear down. This can be ascertained by checking the table in QRH AEP.MISC
One Engine Inoperative – Circling Approach. If affected, plan a FLAP 3 landing
and delay gear extension until level flight is no longer required; a L/G NOT
DOWN warning which will occur if circling below 750ft (can be silenced with the
EMER CANC pb) and a TOO LOW GEAR should be anticipated if below 500ft RA.

\multicite{QRH AEP.MISC}

\section{Bomb on board}

The primary aim is to get the aircraft on the ground and evacuated ASAP.

The secondary aim is to prevent detonation of the device. This is achieved by
preventing further increases in cabin altitude through the use of manual
pressure control and by avoiding sharp manoeuvres and turbulence.

The tertiary aim is to minimise the effect of any explosion. This is achieved by
reducing the diff to 1 psi. The method is to set cabin vertical speed to zero
using manual pressurisation control, then descend to an altitude 2500ft above
cabin altitude. As further descent is required, cabin vertical speed should be
adjusted to maintain the 1 psi diff for as long as possible. Automatic pressure
control is then reinstated on approach. Low speeds reduce the damage from an
explosion but increase the risk of a timed explosion occurring whilst airborne;
a compromise needs to be found. The aircraft should be configured for landing as
early as possible to avoid an explosion damaging landing systems.

In the cabin, procedures are laid down for assessing the risks of moving the
device and for moving the device to the LRBL at door 2R.

\multicite{QRH AER.80, FCOM PRO.AER.MISC}

\section{Stall recovery \emph{(memory item)}}

Aerofoil stall is always and only an angle of attack issue. It is not possible
to directly prove an unstalled condition from attitude and airspeed data. The
flight recorders from the December 2014 Air Asia accident recorded an angle of
attack of 40 degrees (i.e. around 25 degrees greater than critical angle) with
both pitch and roll zero and speeds up to 160kt. Importantly, it is perfectly
possible to be fully stalled in the emergency configurations described in <xref
linkend="unreliable-airspeed"/>. Identification of a fully stalled condition is
thus largely dependent on identifying a high and uncontrollable descent rate
that does not correlate with normal flight path expectations for the attitude
and thrust applied.

To recover from a fully stalled condition, the angle of attack of the aerofoils
must be reduced to below critical. The generic stall recovery is therefore
simply to pitch the nose down sufficiently to break the stall and level the
wings. In normal operations, the velocity vector of the aircraft is around 3°
below the centerline of the aircraft (i.e. an attitude of around 3° is required
to fly level). In a stalled condition, the velocity vector may be 40° or more
below the centerline of the aircraft. Thus the amount of pitch down required to
recover a fully stalled aircraft can be 30° or more.

The aircraft’s thrust vector helps to accelerate the aircraft during the
recovery, and increasing speed along the aircraft’s centerline acts to reduce
the stalled angle of attack. Thus, while thrust is not a primary means of
recovery, it does help. Unfortunately, Airbus have determined that due to the
pitch couple associated with underslung engines, there may be insufficient
longitudinal control authority to pitch the aircraft sufficiently to recover
from a stall if TOGA is selected. It may therefore be necessary to initially
reduce thrust to allow the primary recovery technique to be applied; this is
extremely counterintuitive.

In the two recent Airbus accidents involving stalls, the lack of physical cross
coupling of sidesticks was a major factor. If one pilot elects to hold full back
sidestick, the aircraft cannot be recovered by the other pilot unless the
takeover pushbutton is used. With all the alarms, it is easy to miss “Dual
Input”.

Once there are no longer any indications of the stall, smoothly recover from the
dive, adjust thrust, check speedbrakes retracted and if appropriate (clean and
below 20,000ft) deploy the slats by selecting flaps 1. The load factor
associated with an overly aggressive pull out can induce a secondary stall; on
the flip side, once reattachment of the airflow occurs, drag rapidly diminishes
and exceedance of high speed airframe limitations becomes a threat.

If a stall warner sounds on takeoff it is likely to be spurious since you are
almost certainly in normal law. The procedure in this case is essentially to
initially assume unreliable airspeed and fly TOGA, 15°, wings level until it can
be confirmed that the warning is spurious.

A stall warning may occur at high altitude to indicate that the aircraft is
reaching $\mathrm{\alpha_{buffet}}$. In this case simply reduce the back
pressure on the sidestick and/or reduce bank angle.

\multicite{FCOM PRO.AER.MISC}


\section{Computer reset}

Abnormal computer behaviour can often be stopped by interrupting the power
supply of the affected computer. This can be done either with cockpit controls
or with circuit breakers. The general procedure is to interrupt the power
supply, wait 3 seconds (5 seconds if a C/B was used), restore the power, then
wait another three seconds for the reset to complete. QRH AER.SYSTEM RESET
details the specific procedures for a variety of systems.

On the ground, almost all computers can be reset. MOC can usually supply a reset
procedure if nothing applicable is available in the QRH. The exceptions are the
ECU and EIU while the associated engine is running and the BSCU when the
aircraft is not stopped.

In flight, only the computers listed in the QRH should be considered for reset.

\multicite{QRH AER.SYSTEM RESET}

\section{Landing distance calculations}

Many failures result in a longer than normal landing distance. The QRH inflight
performance section has tables for calculating \V{APP} and Reference Landing
Distances for single failures. These reflect the performance achievable in a
typical operational landing without margin. easyJet requires a factor of 1.15 to
be applied to these distances.

The EFB module provides both factored and unfactored landing distances, and also
can calculate for multiple failures.

The safety factor may be disregarded in exceptional circumstances.

\multicite{QRH IFP, FCOM PER.LDG, EOMB 4.14.2}

\section{Abnormal V Alpha Prot}

If two or more angle of attack vanes become blocked at the same angle during
climb, alpha floor protection will be activated once a Mach number is reached
where the angle of attack at which the vanes were blocked becomes indicative of
an incipient stall condition. Since the flight computer’s attempts to reduce
angle of attack will not be registered by the blocked vanes, a continuous nose
down pitch rate which cannot be stopped with sidestick inputs will result.

Indications of the incipient condition are available from the Alpha Prot and
Alpha max strips. If the Alpha Max strip (solid red) completely hides the Alpha
Prot strip (black and amber) or the Alpha Prot strip moves rapidly by more than
30kt during flight manoeuvres with AP on and speed brakes retracted, blocked
angle of attack vanes should be suspected.

The solution is to force the flight computers into Alternate Law where the
protection does not apply. This is most conveniently done by turning off any two
ADRs. Once in Alternate Law, the stall warning strip (red and black) becomes
available. Since this may be receiving data from a blocked angle of attack vane,
erroneous presentation is possible.

\multicite{OEBPROC-48}

\section{Overspeed Recovery}

In general the response to an overspeed should be to deploy the speedbrake and
monitor the thrust reduction actioned by the autothrust. Disconnection of the
autopilot will not normally be required. If autothrust is not in use, the thrust
levers will need to be manually retarded.

It is possible that the autopilot will automatically disengage and high speed
protection will activate, resulting in an automatic pitch up. In this case,
smoothly adjust pitch attitude as required.

\multicite{FCTM PRO.AER.MISC}

\section{Volcanic Ash Encounter}

Volcanic ash clouds are usually extensive, so a 180° will achieve the quickest exit.

Air quality may be affected, so crew oxygen masks should be donned with 100\%
oxygen to exclude fumes. Passenger oxygen may also need to be deployed.

Be prepared to carry out the unreliable speed procedure as airspeed indications
may be compromised.

Disconnect the autothrust to prevent thrust variations. To minimise the impact
on the engines, if conditions permit thrust should be reduced. Turn on all
anti-ice and set pack flow to high in order to increase bleed demand and thus
increase stall margin. Wing anti-ice should be turned off before restart in case
of double engine flameout.

If engine EGT limits are exceeded, consider precautionary engine shutdown with
restart once clear of volcanic ash. Engine acceleration may be very slow during
restart. Since compressor and turbine blades may have been eroded, avoid sudden
thrust changes.

Damage to the windshield may necessitate an autoland or landing with a sliding
window open.

\multicite{QRH AEP.MISC, FCOM PRO.AEP.MISC, FCTM PRO.AEP.MISC}

\end{document}
